\bf Excuse me, do you have a moment to talk about version control?

### 2017-06-29 README.tex
Various edits + first 3 figures

Why would a statistician use version control, such as
\href{http://git-scm.com}{Git}? And what is the point of hosting your
work online, e.g., on \href{https://github.com}{GitHub}? Could the gains
possibly justify the inevitable pain?

@@

\href{http://git-scm.com}{Git} is a \textbf{version control system}. Its
original purpose was to help groups of developers work collaboratively
on big software projects. Git manages the evolution of a set of files --
called a \textbf{repository} or \textbf{repo} -- in a sane, highly
structured way. It is a bit like the ``Track Changes'' feature from
Microsoft Word, but more rigorous, powerful, and scaled up to multiple
files.

Git has been re-purposed by the data science community
\citep{Ram2013, git-for-humans, ten-simple-rules-git}. We use it to
manage the motley collection of files that make up typical data
analytical projects, which often consist of data, figures, reports, and,
yes, source code. Even those who identify more as statistician than data
scientist generally have a similar mix of files that are the artifacts
of a project.

@@

Many people who don't use Git unwittingly re-invent a poor man's version
of it. They take an important file and distribute it via email. Various
parties make changes, decorating the file name with initials, dates, and
other descriptors. Before you know it, the original file is the root of
a complicated phylogeny that no amount of ``Track changes'' and good
intentions can resolve.

\emph{If you're looking at the PDF draft, you're missing out on some
figures around here in the GitHub Markdown version that hint at the
figure I plan to embed.}

The Git way is to track the evolution of that file, through a series of
commits, each equipped with an explanatory message and a nickname. All
collaborators sync regularly to a common version, acknowledging that the
difficulty of merging only goes up as more changes accumulate
\citep{good-enough}. Especially important versions get a human-readable
tag, to signal a meaningful milestone. Yes, there is some pain in
adopting the formalism of Git, but it is worth it.

\emph{make a figure related to the above (reference those file names?)
but juxtaposing different workflows, inspiration from the Bartlett Git
for Designers talk}

\includegraphics[width=1\linewidth]{bartlett-copies-vs-commits}

@@

Git offers a powerful structure for file management and
\href{https://github.com}{GitHub} complements it with a slick user
interface and distribution mechanism for Git repositories. GitHub is
currently the most popular Git hosting service, providing a home for
Git-based projects on the internet. It is a bit like DropBox or Google
Drive, but more structured, powerful, and programmatic.

The remote host acts as the clearinghouse for a Git-managed project.
This allows others to see the project files, sync up with the current
version, and perhaps even propose or make changes. GitHub has a
well-designed web interface that is a dramatic improvement over
traditional Unix Git servers. Many operations can be done entirely in
the browser, such as viewing current or past versions of a file,
commenting on a recent change, and editing or adding files. GitHub also
offers granular control over who can see, edit, and administer a
project.

\emph{possible images include a modified version of this atlassian image
(no arrows between people, yes arrow from people to github), somehow
connected to a screenshot of a GitHub repo}

\includegraphics[width=1\linewidth]{atlassian-git-tutorials-syncing}

@@

\textbf{GitHub issues} are another powerful feature of the platform.
Recall that we are repurposing Git, a tool designed to facilitate
software development. The issues for a project are its bug tracker. For
projects that are not pure software development, we co-opt this
machinery to organize our to do list more generally. The basic unit is
an issue and you can interact with one in two ways.

@@

Here is an overview of the initial setup process, which happens once or,
for some steps, once per computer. This is excerpted from
\href{http://happygitwithr.com}{Happy Git and GitHub for the useR},
which holds battle-tested instructions honed over several years in
\href{http://stat545.com}{STAT 545} at the University of British
Columbia.

\begin{itemize}
\item
  Register for a free account with GitHub.
\item
  Install Git. Depending on your OS, Git might already be installed. But
  many may need to install it or will choose to update to a more recent
  version. Some basic configuration is critical, such as setting your
  username and email.
\item
  Install a local Git client \emph{(optional but highly recommended}). A
  Git client is software that provides a graphical user interface for
  Git, which is otherwise command-line only. If you are an R user, you
  will find that
  \href{https://www.rstudio.com/products/rstudio-desktop/}{RStudio}
  provides a great deal of this functionality. There are some notable
  gaps, however, so you might still choose to install a dedicated and
  comprehensive Git client such as
  \href{https://www.sourcetreeapp.com}{SourceTree} or
  \href{https://www.gitkraken.com}{GitKraken}. Git is a file-based
  system, so you can do some operations from RStudio, others from
  SourceTree, and others from the shell.
\item
  Confirm, with a practice repository, that local Git can send and
  receive the current version of the repository on GitHub, known as
  \textbf{pushing} and \textbf{pulling}, respectively.
\end{itemize}

@@

\emph{this is really crying out for an example and/or diagrams,
something that shows commits unfolding and let's you illustrate a diff,
sort of like this from the barlett talk}

\includegraphics[width=1\linewidth]{bartlett-commit-history}

Consider version A of a file and a modified version, version B. Assume
that version A was part of a Git commit and version B was part of the
next commit. The set of differences between A and B is called a ``diff''
and Git users contemplate diffs a lot. Diff inspection is how you
re-explain to yourself how version A differs from version B. Diff
inspection is not limited to adjacent commits. You can inspect the diffs
between any two commits.

In fact, Git's notion of version B of your file is as an accumulation of
diffs. At some earlier point, the file was created in the first place.
That version of the file was part of a commit and, therefore, a diff.
Git stores Version A of the file as the initial version, plus all the
intervening diffs in the history that affect the file. And version B
simply includes one more. We'll set these internal details aside now,
but understanding the importance of these deltas will eventually make
Git's operations less baffling.

@@

Every time you make a commit you must also write a short \textbf{commit
message}. Ideally, this conveys the motivation for the change. Remember,
the diff will show the content. When you revisit a project after a break
or need to digest recent changes made by a colleague, looking at the
\textbf{history}, by reading commit messages and skimming through diffs,
is an extremely efficient way to get up to speed.

Every commit needs some sort of nickname, so you can identify it. Git
does this automatically, assigning each commit what is called a SHA, a
seemingly random string of 40 letters and numbers (it is not, in fact,
random but is a SHA-1 checksum hash of the commit). Though you will be
exposed to these, you don't have to handle them directly very often and,
when you do, usually the first 7 characters suffice. You can also
designate certain snapshots as special with a \textbf{tag}, which is a
name of your choosing. In a software project, it is typical to tag a
release with its version, e.g., ``v1.0.3''. For a manuscript or
analytical project, you might tag the version submitted to a journal or
transmitted to external collaborators.

@@

This may seem unrelated to Git, GitHub, and R, but we need to talk about
\href{https://daringfireball.net/projects/markdown/syntax}{Markdown}.
Markdown is a markup language, like HTML and LaTeX, but designed to be
as lightweight as possible. The goal is still to separate form and
content, but also to prioritize human-readability, even at the cost of
fancy features. Markdown is in wide use on sites like
\href{https://en.support.wordpress.com/markdown/}{WordPress},
\href{https://stackoverflow.com/editing-help}{StackOverflow}, and, yes,
\href{https://help.github.com/categories/writing-on-github/}{GitHub}.
These sites use Markdown because it allows a diverse population of site
users to author decent-looking web content, with hyperlinks and some
formatting. Do not build this up into some heroic, LaTeX-level learning
task, for it is not. If you can write an email, you can write Markdown.

@@

Markdown is extra special for R users because of
\href{http://rmarkdown.rstudio.com}{R Markdown}, which is just Markdown
that includes chunks of R code. Again, do not regard R Markdown as
something you must clear your schedule to learn. If you can write email
and a bit of R code, you can write R Markdown. The
\href{https://CRAN.R-project.org/package=rmarkdown}{rmarkdown package}
\citep{rmd-pkg} converts R Markdown (\texttt{.Rmd} files) to Markdown
(\texttt{.md} files), running the code and inserting the results,
including figures, into the document. This is powered by another
package, \href{https://CRAN.R-project.org/package=knitr}{knitr}
\citep{knitr-pkg, knitr-book}, under the hood. This process is made
especially easy in RStudio, but is by no means limited to users of that
application.

@@

You do not, in fact, even need to work in R Markdown to exploit this. It
\href{http://rmarkdown.rstudio.com/articles_report_from_r_script.html}{works
with plain R scripts} as well. You can use exactly the same machinery to
prepare a rendered version of an R script, i.e.~to go from \texttt{.R}
to \texttt{.md}. Again, RStudio makes this especially easy, but this is
not limited to RStudio. Once the Markdown file is pushed to GitHub, it
is as if the reader has run your code or is able to look over your
shoulder at your R session. This provides an lightweight system for
exposing work-in-progress to collaborators, without slowing down to
create separate reports. Comment lines that begin with
\texttt{\#\textquotesingle{}} are elevated to top-level prose, providing
a way to make the document more welcoming for a reader. Once there are
many prose comments, you might decide to switch from \texttt{.R} to
\texttt{.Rmd}, have proper top-level prose, and move the code down into
chunks.

@@

Let's zoom back out again and consider R Markdown and the rmarkdown
package more generally. I want to point out that R Markdown can be
rendered to many more formats than Markdown! I have emphasized the
production of Markdown here because it is extremely useful in
GitHub-hosted projects and that seems to be underappreciated. But the
rmarkdown package can convert \texttt{.Rmd} to a wide array of
\href{http://rmarkdown.rstudio.com/lesson-9.html}{output formats},
including HTML, PDF, and Word (\texttt{.docx}). The Markdown must be
created regardless. Sometimes it is all you need, such as on GitHub. In
other contexts, it is just a necessary intermediate and may even be
discarded.

@@

The files in a project arise in different ways and play different roles.
A critical issue for workflow happiness is to figure out how you want to
handle different file types with respect to Git. You can direct Git to
ignore specific files or file types, such as autosaves created by your
editor. This reduces noise and clutter: Git will not pester you to
commit changes to these files and they will not appear in your GitHub
repository. A file that Git does not ignore is said to be
\textbf{tracked}. How should you think about this?

@@

\textbf{Is it useful to someone?} If so, track and share! There is a
taboo against committing derived products, inherited from Git's software
development roots, because their typical product is a platform-specific
executable. This rationale, however, does not apply to many data science
products. Rendered reports, figures, and cleaned data are often
extremely valuable to others. Make them readily available.

@@

\begin{itemize}
\tightlist
\item
  Some derived files are too miserable to read casually, such as
  \texttt{.csv} files of processed results or \texttt{.html} derived
  from \texttt{.Rmd}, but are still worth tracking with Git. When you
  re-run an analysis with updated input data or after updating R
  packages, \emph{the diffs are often quite modest} and help you
  pinpoint unexpected changes.
\item
  GitHub has excellent
  \href{https://help.github.com/categories/working-with-non-code-files/}{support
  for a variety of non-code files}, such as CSV and TSV. It also
  \href{https://help.github.com/articles/rendering-and-diffing-images/}{displays
  and provides visual diffs} for the most common image formats, which is
  extremely useful for spotting unexpected changes in figures.
\item
  Never, ever forget that Markdown is highly privileged on GitHub and
  is, therefore, a crucial output format for rendered \texttt{.R} and
  \texttt{.Rmd}.
\end{itemize}

\textbf{Will it cause problems with Git/GitHub?} This boils down to
whether Git diffs are informative, whether GitHub can render the file
usefully, and the file's likely effect on Git operations. A file that is
large and changing often can make your repository bloated and slow down
pushes and pulls. If a file is binary, such as a Word document or Excel
spreadsheet, you will not get human-readable diffs, nor can GitHub
display it in the browser. Html is not rendered on GitHub, but is,
instead, displayed in raw form. Binary files are also a prime source of
merge conflicts (see below), because they are beyond the reach of Git's
sophisticated automatic merging logic. This all implies that adoption of
Git/GitHub should prompt a pivot away from \texttt{.docx} and
\texttt{.pdf} as primary output formats and towards \texttt{.md}, at
least during periods of rapid development.

@@

Collaboration is the most compelling reason to manage a project with Git
and GitHub. My definition of collaboration includes hands-on
participation by multiple people as well as an asymmetric model, in
which some people are active makers and others only read or review.

Consider two different models of collaboration on a document:

\begin{itemize}
\item
  \textbf{Edit, save, attach.} In this workflow, everyone has one (or
  more!) copies of the document and they circulate via email attachment,
  accumulating initials and dates in the filename. Which one is
  ``master''? Does this question even make sense anymore? How do
  different versions of the document relate to each other? If you want
  to see a version combining the best versions of each section, how
  would you reconcile the different copies into one? All of this usually
  gets sorted out by social contract, a fairly manual process, and at
  least one miserable person.
\item
  \textbf{Google Doc.} In this workflow, there is only one copy of the
  document and it lives in the cloud. Anyone can access the most recent
  version on demand. Anyone can edit or comment or propose a change and
  this is immediately available to everyone else. Anyone can see who's
  been editing the document and how and, if disaster strikes, can revert
  to a previous version. A great deal of ambiguity and annoying
  reconciliation work has been designed away.
\end{itemize}

Managing a project via Git/GitHub is much more like the Google Doc
scenario, but also offers some of the attractive features of ``edit,
save, attach''. With Git/GitHub, collaborators can work offline and
there can be independent lines of development. The real power comes from
regular and structured reconciliation of all versions of all the files
in the project via Git/GitHub. It is definitely more complicated than
collaborating on a Google Doc, but also more powerful.

@@

Git is a decentralized version control system, meaning each collaborator
has their own complete copy of the repo and its history. Everyone can
work offline and/or simultaneously, but with regular syncing to GitHub.
GitHub plays the role of another collaborator, but a very special one.
By convention, everyone agrees that GitHub keeps the master copy of the
project. GitHub is the clearinghouse. The joke is that GitHub puts the
``central'' in decentralized version control. You pull regularly from
GitHub, to receive and integrate changes made by your collaborators. You
also push regularly to GitHub, to return the favor, and to maintain its
status as the comprehensive, authoritative version of the project.

\emph{These two points were originally written as comparisons with
Google Docs, but now there are intervening paragraphs. Is it too
disconnected? Is it worth weaving in?}

Merge conflicts are the most frustrating thing about using Git and
GitHub. You can avoid them if you only work alone, on one computer, but
I've also said that collaboration is the best reason to use GitHub! So
this problem must be confronted.

What is a merge conflict? Here is a typical first encounter: your
collaborator makes a change to a file, commits it locally, and pushes to
GitHub. Meanwhile you also make a different change to the same file and
also commit locally. When you try to push your commit to GitHub, it will
fail because there are commits on GitHub that you do not have. You must
pull from GitHub. The good news is that quite often, this will ``just
work'', i.e.~the GitHub version and your version will merge cleanly. Git
is quite clever at reconciliation and changes to different files or even
distinct parts of the same file will merge. If this pull and merge goes
smoothly, you'll able to push your changes and the cycle goes on.

But sometimes it's not clear how to reconcile your changes with the new
ones from GitHub and you get a \emph{merge conflict}. You must inspect
the locations of conflict, which Git marks for you. You will pick one
version or the other -- or create a hybrid -- at each location of
conflict and mark it as resolved. Once you've resolved all conflicts,
you will be able to push a version integrating your recent changes to
GitHub.

The best way to deal with merge conflicts is to avoid them altogether.
This is another reason for all parties to commit, pull, and push often.
Small changes, integrated frequently, in non-binary files, are the
easiest for Git to automatically merge for you. The difficulty of
merging (by Git or by you) is proportional to the evolutionary distance
between two lines of work. The presence of frequently-changing binary
files also increases the burden. So make lots of small commits, sync
regularly with GitHub, and only track binary files with good reason.

\emph{once I understand Nick's point about binary files causing merge
conflicts
\url{https://github.com/dsscollection/git-github-for-stats/issues/6} I
may need to reword the above}

@@

GitHub offers several ways to host a proper website directly from a
repository, collectively known as
\href{https://help.github.com/categories/github-pages-basics/}{GitHub
Pages}. At one extreme, as long as you've got one Markdown file,
\href{https://github.com/blog/2289-publishing-with-github-pages-now-as-easy-as-1-2-3}{GitHub
Pages can create a simple website} automatically. At the other extreme,
sophisticated users can take full advantage of the
\href{https://jekyllrb.com}{Jekyll static site generator}. The STAT 545
website (\href{http://stat545.com}{stat545.com},
\href{https://github.com/STAT545-UBC/STAT545-UBC.github.io}{GitHub
repo}) falls on the more primitive side.
\href{http://rmarkdown.rstudio.com/rmarkdown_websites.html}{R Markdown
websites}, \href{https://bookdown.org}{bookdown}
\citep{bookdown-pkg, bookdown-book}, and
\href{https://bookdown.org/yihui/blogdown/}{blogdown} provide several
R-focused options for rendering the pages \emph{en masse}.

@@

\textbf{Use conventional file extensions.} GitHub is very code-focused
and will apply proper syntax high-lighting for almost any language you
can think of, if you use one of the standard file extensions. This also
has advantages for people searching GitHub and trying to filter by
language.

@@

>>>>

Why would a statistician use a version control system, such as
\href{http://git-scm.com}{Git} \citep{git}? And what is the point of
hosting your work online, e.g., on \href{https://github.com}{GitHub}
\citep{github}? Could the gains possibly justify the inevitable pain?

@@

\href{http://git-scm.com}{Git} is a \textbf{version control system}. Its
original purpose was to help groups of developers work collaboratively
on big software projects. Git manages the evolution of a set of files --
called a \textbf{repository} or \textbf{repo} -- in a sane, highly
structured way. It is like the ``Track Changes'' feature from Microsoft
Word, but more rigorous, powerful, and scaled up to multiple files.

Git has been re-purposed by the data science community
\citep{Ram2013, git-for-humans, ten-simple-rules-git}. We use it to
manage the motley collection of files that make up typical data
analytical projects, which consist of data, figures, reports, and, yes,
source code. Even those who identify more as statistician than data
scientist generally have a similar mix of files that are the artifacts
of a project.

@@

Many people who don't use Git unwittingly re-invent a poor man's version
of it. Figure X depicts a hypothetical analysis of the iris data,
captured in a single R source file. With informal version control,
contributors create derivative copies of \texttt{iris.R}, decorating the
file name with initials, dates, and other descriptors. Even when working
alone, this leads to multiple versions of \texttt{iris.R} of
indeterminate relatedness (Fig Xa). In collaborative settings based on
email distribution, the original file swiftly becomes the root of a
complicated phylogeny that no amount of ``Track changes'' and good
intentions can resolve (Fig Xb).

\includegraphics[width=1\linewidth]{diy-vs-git-solo-vs-collab}

The Git way is to track the evolution of \texttt{iris.R}, through a
series of commits, each equipped with an explanatory message and a
nickname. Figure Xc depicts this sort of linear \emph{in situ}
development. Figure Xd shows the same history for a common collaborative
Git workflow, where contributors work independently but sync regularly
to a common version. Especially important versions get a human-readable
tag, to signal a meaningful milestone. Yes, there is some pain in
adopting the formalism of Git, but it is worth it.

@@

We've explored Git's powerful structure for file management, so where
does GitHub fit in? \href{https://github.com}{GitHub} complements Git by
providing a slick user interface and distribution mechanism for Git
repositories. Git is the software you will use locally to record changes
to a set of files. GitHub is a hosting service that provides a Git-aware
home for such projects on the internet. It is like DropBox or Google
Drive, but more structured, powerful, and programmatic.

The remote host acts as the clearinghouse for a Git-managed project.
This allows others to browse project files, explore their history, sync
up with the current version, and perhaps even propose or make changes.
GitHub's well-designed web interface is a dramatic improvement over
traditional Unix Git servers. Many operations can be done entirely in
the browser, including editing or adding files. It is easy to create a
hyperlink to a specific file or location in a file, at a specific
version, which can make meta-conversations about project code or reports
much more productive. GitHub also offers granular control over who can
see, edit, and administer a project.

\includegraphics[width=1\linewidth]{your-repo-their-repo-central-remote-repo}

@@

\textbf{GitHub issues} are another powerful feature of the platform.
Recall that we are repurposing Git, a tool that facilitates software
development. Think of the issues for a project as its bug tracker. For
projects that are not pure software development, we co-opt this
machinery to organize our to-do list more generally. The basic unit is
an issue and you can interact with one in two ways.

@@

If I've convinced you to experiment with Git and GitHub, you need to do
some initial setup. These first steps happen one or, for some steps,
once per computer. This is excerpted from
\href{http://happygitwithr.com}{Happy Git and GitHub for the useR},
which holds battle-tested instructions honed over several years in
\href{http://stat545.com}{STAT 545} at the University of British
Columbia.

\begin{itemize}
\item
  Register for a free account with GitHub.
\item
  Install Git. Depending on your OS, Git might already be installed. But
  many will need to install it or will choose to update to a more recent
  version. Some basic configuration is critical, such as setting your
  username and email.
\item
  Install a local Git client, \emph{optional but highly recommended}. A
  Git client provides a graphical user interface for Git, which is
  otherwise command-line only. If you are an R user, you will find that
  \href{https://www.rstudio.com/products/rstudio-desktop/}{RStudio}
  provides a great deal of this functionality. There are some notable
  gaps, however, so you might still choose to install a dedicated and
  comprehensive Git client such as
  \href{https://www.sourcetreeapp.com}{SourceTree} or
  \href{https://www.gitkraken.com}{GitKraken}. Git just operates on
  files, so you can do some operations from RStudio, others from
  SourceTree, and others from the shell.
\item
  Confirm, with a practice repository, that local Git can send and
  receive the current version of the repository on GitHub, known as
  \textbf{pushing} and \textbf{pulling}, respectively.
\end{itemize}

@@

\includegraphics[width=1\linewidth]{commit-diff-sha-tag}

Figure X is another look at our fictional analysis of the iris data,
focusing on the evolution of its script \texttt{iris.R}. Consider
version A of this file and a modified version, version B. Assume that
version A was part of one Git commit and version B was part of the next
commit. The set of differences between A and B is called a ``diff'' and
Git users contemplate diffs a lot. Diff inspection is how you re-explain
to yourself how version A differs from version B. Diff inspection is not
limited to adjacent commits. You can inspect the diffs between any two
commits.

In fact, Git's notion of any specific version of \texttt{iris.R} is as
an accumulation of diffs. If you go back far enough, you find the commit
where the file was created in the first place. Every later version is
stored by Git as that initial version, plus all the intervening diffs in
the history that affect the file. We'll set these internal details aside
now, but understanding the importance of these deltas will make Git's
operations less baffling in the long run.

@@

Every time you make a commit you must also write a short \textbf{commit
message}. Ideally, this conveys the motivation for the change. Remember,
the diff will show the content. When you revisit a project after a break
or need to digest recent changes made by a colleague, looking at the
\textbf{history}, by reading commit messages and skimming through diffs,
is an extremely efficient way to get up to speed. Figure X shows the
messages associated with the last three commits.

Every commit needs some sort of nickname, so you can identify it. Git
does this automatically, assigning each commit what is called a SHA, a
seemingly random string of 40 letters and numbers (it is not, in fact,
random but is a SHA-1 checksum hash of the commit). Though you will be
exposed to these, you don't have to handle them directly very often and,
when you do, usually the first 7 characters suffice. The commit messages
in Figure X are prefixed by such abbreviated SHAs. You can also
designate certain snapshots as special with a \textbf{tag}, which is a
name of your choosing. In a software project, it is typical to tag a
release with its version, e.g., ``v1.0.3''. For a manuscript or
analytical project, you might tag the version submitted to a journal or
transmitted to external collaborators. Figure X shows a tag,
``draft-01'', associated with the last commit.

@@

This may seem unrelated to Git, GitHub, and R, but it is now necessary
to talk about
\href{https://daringfireball.net/projects/markdown/syntax}{Markdown}.
Markdown is a markup language, like HTML and LaTeX, but designed to be
as lightweight as possible. The goal is still to separate form and
content, but also to prioritize human-readability, even at the cost of
fancy features. Markdown is in wide use on sites like
\href{https://en.support.wordpress.com/markdown/}{WordPress},
\href{https://stackoverflow.com/editing-help}{StackOverflow}, and, yes,
\href{https://help.github.com/categories/writing-on-github/}{GitHub}.
These sites use Markdown because it allows a diverse population of site
users to author decent-looking web content, with hyperlinks and some
formatting. Do not build this up into some heroic, LaTeX-level learning
task, for it is not. If you can write an email, you can write Markdown.

@@

Markdown is especially holy for R users because of
\href{http://rmarkdown.rstudio.com}{R Markdown}, which is just Markdown
that includes chunks of R code. Again, do not regard R Markdown as
something you must clear your schedule to learn. If you can write email
and a bit of R code, you can write R Markdown. The
\href{https://CRAN.R-project.org/package=rmarkdown}{rmarkdown package}
\citep{rmd-pkg} converts R Markdown (\texttt{.Rmd} files) to Markdown
(\texttt{.md} files), running the code and inserting the results,
including figures, into the document. This is powered by another
package, \href{https://CRAN.R-project.org/package=knitr}{knitr}
\citep{knitr-pkg, knitr-book}, under the hood. This process is made
especially easy in RStudio, but is by no means limited to users of that
application.

@@

You do not, in fact, even need to work in R Markdown to exploit this. It
\href{http://rmarkdown.rstudio.com/articles_report_from_r_script.html}{works
with plain R scripts} as well. You can use exactly the same machinery to
prepare a rendered version of an R script, i.e.~to go from \texttt{.R}
to \texttt{.md}. Again, RStudio makes this especially easy, but this is
not limited to RStudio. Once the Markdown file is pushed to GitHub, it
is as if the reader has run your code or is able to look over your
shoulder at your R session. This provides a lightweight system for
exposing work-in-progress to collaborators, without slowing down to
create separate reports. Comment lines that begin with
\texttt{\#\textquotesingle{}} are elevated to top-level prose, providing
a way to make the document more welcoming for a reader. Once there are
many prose comments, you might decide to switch from \texttt{.R} to
\texttt{.Rmd}, have proper top-level prose, and move the code down into
chunks.

@@

Before we move on, I want to zoom out and revisit R Markdown and the
rmarkdown package more generally. Note that R Markdown can be rendered
to many more formats than Markdown, including HTML, PDF, and Microsoft
Word, and can incorporate code chunks in many languages other than R.
Markdown is emphasized here, because it is extraordinarily useful in
GitHub-hosted projects. Even when targetting another output format, the
Markdown must be created as a necessary intermediate. Sometimes it is
all you neeed, such as on GitHub, and in other contexts it may even be
discarded.

@@

The files in a project arise in different ways and play different roles.
A critical issue for workflow happiness is to figure out how you want to
handle different file types with respect to Git. You can direct Git to
ignore specific files or file types, such as autosaves created by your
editor. This reduces noise and clutter: Git will not pester you to
commit changes to these files and they will not appear in your GitHub
repository. A file that Git does not ignore is said to be
\textbf{tracked}. Here's a useful framework for thinking about what to
track:

@@

\textbf{Is it useful to someone?} If so, track and share! There is a
taboo against committing derived products, inherited from Git's software
development roots, because the typical product in that context is a
platform-specific executable. This rationale, however, does not apply to
many data science products. Rendered reports, figures, and cleaned data
are often extremely valuable to others. Make them readily available.

@@

\begin{itemize}
\tightlist
\item
  Some derived files are too miserable to read casually, such as
  \texttt{.csv} files of processed results or \texttt{.html} derived
  from \texttt{.Rmd}, yet they are still worth tracking with Git. When
  you re-run an analysis with updated input data or after updating R
  packages, \emph{the diffs are often quite modest} and help you
  pinpoint unexpected changes.
\item
  GitHub has excellent
  \href{https://help.github.com/categories/working-with-non-code-files/}{support
  for a variety of non-code files}, such as CSV and TSV. It also
  \href{https://help.github.com/articles/rendering-and-diffing-images/}{displays
  and provides visual diffs} for the most common image formats, which is
  extremely useful for spotting unexpected changes in figures.
\end{itemize}

\textbf{Will it actively cause problems with Git/GitHub?} This boils
down to the file's likely effect on Git operations. A file that is large
and changing often can make your repository bloated and slow down pushes
and pulls. If a file is binary, such as a Word document or Excel
spreadsheet, you will not get human-readable diffs anyway, nor can
GitHub display the content in the browser. Binary files are also a
reliable source of merge conflicts (see below), because they are beyond
the reach of Git's sophisticated automatic merging logic. A large binary
file that changes often is, therefore, the worst of all worlds. This
implies that adoption of Git/GitHub rewards a pivot away from
\texttt{.docx}, \texttt{.xlsx}, and \texttt{.pdf} as primary file
formats and towards \texttt{.Rmd}, \texttt{.md}, and \texttt{.csv}, at
least during periods of rapid development.

@@

Collaboration is the most compelling reason to manage a project with Git
and GitHub. My definition of collaboration includes hands-on
participation by multiple people, including your past and future self,
as well as an asymmetric model, in which some people are active makers
and others only read or review.

Consider two different ways to collaborate on a document:

\begin{itemize}
\item
  \textbf{Edit, save, attach.} In this workflow, everyone has one (or
  more!) copies of the document, which circulate as email attachments,
  accumulating initials and dates in the filename. Which one is
  ``master''? Does this question even make sense anymore?!? If you want
  a version combining the edits made by different authors to different
  sections, how do you reconcile the copies into one? All of this
  usually gets sorted out by social contract, a fairly manual process,
  and at least one miserable person.
\item
  \textbf{Google Doc.} In this workflow, there is only one copy of the
  document and it lives in the cloud. Anyone can access the most recent
  version on demand. Anyone can edit or comment or propose a change and
  this is immediately available to everyone else. Anyone can see who's
  been editing the document and how and, if disaster strikes, can revert
  to a previous version. A great deal of ambiguity and annoying
  reconciliation work has been designed away.
\end{itemize}

Managing a project via Git/GitHub is much more like the Google Doc
scenario, but also offers some of the attractive features of ``edit,
save, attach''. With Git/GitHub, collaborators can work offline and
there can be independent lines of development, i.e.~branches. The killer
feature is that Git/GitHub enable regular and structured reconciliation
of all versions of all the files in the project. It is definitely more
complicated than collaborating on a Google Doc, but also more powerful.

@@

Git is a decentralized version control system, meaning each collaborator
has their own complete copy of the repo and its history. Everyone can
work offline and/or simultaneously. GitHub plays the role of another
collaborator, but a very special one. By convention, everyone agrees
that GitHub is the clearinghouse, i.e.~it holds the master copy of the
project. The joke is that GitHub puts the ``central'' in decentralized
version control. You pull regularly from GitHub, to receive and
integrate changes made by your collaborators. You also push regularly to
GitHub, to return the favor, and to maintain its status as the
comprehensive, authoritative version of the project.

What if two people have made changes to the repository? Imagine that
your collaborator makes a change to a file, commits it locally, and
pushes to GitHub. Meanwhile you also make a different change to the same
file and also commit locally. When you try to push your commit to
GitHub, you will fail because there are commits on GitHub that you do
not have. You must pull from GitHub. The good news is that quite often,
this will ``just work'', i.e.~the GitHub version and your version will
merge cleanly. Git is quite clever at reconciliation and changes to
different files or even distinct parts of the same file will merge. This
derives from the ``diff'' based model of Git described earlier. After a
successful merge, you can push your changes and the cycle goes on.

But sometimes it's not clear how to reconcile your changes with the new
ones from GitHub and you get a \emph{merge conflict}. Merge conflicts
are the most frustrating thing about using Git and GitHub. You can avoid
them if you only work alone, on one computer, but I've also said that
collaboration is the best reason to use GitHub! So this problem must be
confronted.

What is a merge conflict? It happens when Git can't be certain how to
jointly apply the diffs from two different commits to their common
parent. At each location of conflict, you must pick one version or the
other -- or create a hybrid -- and mark it as resolved. Many Git clients
have special tooling for this specific task, which can be very
convenient. Once you've resolved all conflicts, you will be able to
finalize the merge and push a version integrating your recent changes to
GitHub.

The best way to deal with merge conflicts is to prevent them. This is
another reason for all parties to commit, pull, and push often. Small
changes, integrated frequently, in non-binary files, are the easiest for
Git to automatically merge for you. The difficulty of merging (by Git or
by you) is proportional to the evolutionary distance between two lines
of work. The presence of frequently-changing binary files also increases
the burden. So make lots of small commits, sync regularly with GitHub,
and only track binary files with good reason.

@@

GitHub also offers several ways to host a proper website directly from a
repository, collectively known as
\href{https://help.github.com/categories/github-pages-basics/}{GitHub
Pages}. At one extreme, as long as you've got one Markdown file,
\href{https://github.com/blog/2289-publishing-with-github-pages-now-as-easy-as-1-2-3}{GitHub
Pages can create a simple website} automatically. At the other extreme,
sophisticated users can take full advantage of the
\href{https://jekyllrb.com}{Jekyll static site generator}. The STAT 545
website (\href{http://stat545.com}{stat545.com},
\href{https://github.com/STAT545-UBC/STAT545-UBC.github.io}{GitHub
repo}) falls on the more primitive side.
\href{http://rmarkdown.rstudio.com/rmarkdown_websites.html}{R Markdown
websites}, \href{https://bookdown.org}{bookdown}
\citep{bookdown-pkg, bookdown-book}, and
\href{https://bookdown.org/yihui/blogdown/}{blogdown} provide several
R-focused options for rendering the pages \emph{en masse}.

@@

\textbf{Use conventional file extensions.} GitHub is very code-aware and
will apply proper syntax highlighting for almost any language you can
think of, if you use one of the standard file extensions. This also has
advantages for people searching GitHub and trying to filter by language.

@@

\url{http://mynameismichelle.com/git-frost/}

\emph{These two points were originally written as comparisons with
Google Docs, but now there are intervening paragraphs. I think I will
cut them.}
